A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-28 text/survival.Rmd
cut all text to conform to Cell's 45k character limit

---
output:
  word_document: default
  html_document: default
---
If our molecular pathway disruption clusters represent biological states
distinct from tissue of origin, they may have different prognoses within cancer
types or across all cancers. These analyses are limited by confounding factors
of age and stage at diagnosis, sex, ethnicity, and tissue-specific disease
progression. To explore these ideas, we used Bayesian inference to test models
(see Methods and **Supplemental Methods** for greater detail) of survival using
public longevity data from the CDC, accounting for age, gender and ethnicity.
Our model estimates the effect of cancer type and cluster-specific cancer
effects independently, resulting in a cancer and cluster-specific estimate of
the aging rate multiplier, $k$.

To understand the model, we show in **Figure 7A** simulated survival time curves
for four related groups. We start with a cohort of women aged 30 and show their
expected survival (light blue). By setting their cancer rate multiplier to $k=2$
we can simulate the effect of a moderately deadly cancer (yellow). Compare this
to a randomly selected group of women ranging in age from 30 to 70 years old
(dark blue). Immediately the survival curve changes due to the mixture of ages,
without any malignancy. Adding a $k=2$ malignancy further reduces expected
lifespan (red). We can see that for different tissues, simply due to the change
in distribution of age-at-diagnosis, we should expect equally deadly cancers to
have *different* survival curves. The model takes this into account and the
aggressiveness of the cancer can be estimated accurately without the confounding
effect of age.

@@

**Figure 7: Comparison of different cancers, simulated and actual results
results.** To illustrate the risk model, we compare survival of four simulated
cohorts. A) A group consisting of randomly selected 30 year old
women (light blue) vs. an equal sized group of 30 year old women with cancer, k =
2 (light orange). To illustrate how the mix of ages alters the survival curve,
we compare a randomly selected group of people with ages uniformly distributed
from 30 to 70 (dark blue) and a randomly selected group with uniformly
distributed ages and cancer k = 2 (dark orange). B,C,D,E) We compare a randomly
selected group with correct age distribution for the given cancer (turqoise) to
the model predicted survival for the given average k (blue) and the actual
survival for the patients in our dataset (orange). As can be seen, Stomach
Cancer (STAD), Glioblastoma (GBM), and Pancreatic Cancer (PAAD) are all very
deadly. The modeled results fit the general trend for the actual patients.
Thyroid Cancer (THCA) on the other hand has very little if any effect on our
estimates of life expectancy, in fact there may be a slight benefit which may be
associated with a bias in diagnosis towards patients who are more health
conscious, or higher socioeconomic status, with more access to care compared to
random population members.

@@

We found that cancer types, as expected, have a range of prognoses relative to
the general population. For example in **Figure 7B-E** we see three particularly
deadly cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one
cancer where diagnosis apparently decreases risk relative to background
(Thyroid: THCA). Cancers with posterior probability for relative risk of less
than 1 such as THCA should be interpreted carefully. This Bayesian model is a
model for a state of information. The information that a person is diagnosed
with cancer may lead us to expect that they will live a shorter time than the
general population of matched age (k > 1), or a longer time than the general
population of matched age (k < 1). One mechanism for a shorter result is that
the cancer is aggressive and we can expect it to rapidly injure the body,
causing death. One explanation for longer results may be that the cancer is
relatively mild, and therefore diagnosis is potentially an informational signal
that the patient is health conscious, with the comparison group having more
people whose cancers go undiagnosed. It's important to note that the diagnosis
can increase our expectation of life relative to the comparison group, even if
it decreases the expectation of life of the individual relative to the
counterfactual where they did not have cancer.

>>>>

---
output:
  word_document: default
  html_document: default
---
If our clusters represent biological states
distinct from tissue of origin, they may have different prognoses within cancer
types or across all cancers. These analyses are limited by confounding factors
of age and stage at diagnosis, sex, ethnicity, and tissue-specific disease
progression. To explore these ideas, we used Bayesian inference to test models
of survival using
public longevity data from the CDC.
Our model estimates the effect of cancer type and cluster-specific cancer
effects independently, resulting in a cancer and cluster-specific estimate of
the aging rate multiplier, $k$.

To illustrate the model, in **Figure 7A** we simulate survival time curves
for four related groups. We start with a cohort of women aged 30 and show their
expected survival (light blue). By setting their cancer rate multiplier to $k=2$
we can simulate the effect of a moderately deadly cancer (yellow). Compare this
to a randomly selected group of women ranging in age from 30 to 70 years old
(dark blue). Immediately the survival curve changes due to the mixture of ages,
without any malignancy. Adding a $k=2$ malignancy further reduces expected
lifespan (red). We can see that for different tissues, simply due to the change
in distribution of age-at-diagnosis, we should expect equally deadly cancers to
have *different* survival curves. The model takes this into account and the
aggressiveness of the cancer can be estimated accurately without the confounding
effect of age.

@@

**Figure 7: Comparison of different cancers, simulated and actual results
results.** To illustrate the risk model, we compare survival of four simulated
cohorts. A) A group consisting of randomly selected 30 year old women (light
blue) vs. an equal sized group of 30 year old women with cancer, k = 2 (light
orange). To illustrate how the mix of ages alters the survival curve, we compare
a randomly selected group of people with ages uniformly distributed from 30 to
70 (dark blue) and a randomly selected group with uniformly distributed ages and
cancer k = 2 (dark orange). B,C,D,E) We compare a randomly selected group with
correct age distribution for the given cancer (turqoise) to the model predicted
survival for the given average k (blue) and the actual survival for the patients
in our dataset (orange). Differences with actual data are likely due to a
mixture of cluster identities which we exclude from these simulations for
simplicity.

@@

We found that cancer types, as expected, have a range of prognoses relative to
the general population. In **Figure 7B-E** we see three particularly deadly
cancers (Stomach:STAD, Glioblastoma:GBM, and Pancreatic:PAAD), and one cancer
where diagnosis apparently decreases risk relative to background (Thyroid:
THCA). Cancers with posterior probability for relative risk of less than 1 such
as THCA should be interpreted carefully. This Bayesian model is a model for a
state of information. The information that a person is diagnosed with cancer may
lead us to expect that they will live a shorter time than the general population
of matched age (k > 1), or a longer time than the general population of matched
age (k < 1). A shorter prognosis could result from the cancer being aggressive,
injuring the body and causing death. By contrast, a longer prognosis may result
from the cancer being relatively mild, and therefore diagnosis could be an
informational signal that the patient is health conscious, with the comparison
group having more people whose cancers go undiagnosed. It is important to note
therefore that the diagnosis can increase our expectation of life relative to
the comparison group, even if it decreases the expectation of life of the
individual relative to the counterfactual where they did not have cancer.
